Yeast RNA-seq 48 sample simulation study

### 2018-07-17 datasets/RNA-Seq/polyester-simulation.Rmd
Use the same dataset for informative and uninformative covariates within replicate

```{r}
#' @param counts is the full count dataset
#' @param paramPolyester is the parameters from running polyester. 
#' Contains information on the mean-variance relationship used to 
#' simulate RNA-Seq counts. 
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
simulateOneSplit <- function(X, counts, paramPolyester, 
                             nDE, sampleSize, bd){

@@

    # randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
    log2FC <- rep(0, nrow(counts))
    # log2FC[DE] <- rtruncnorm(nDE, 1, sd=.5) * sample(c(1,-1), size=nDE, replace=TRUE)
    log2FC[DE] <- c(rnorm(nDE, sd = 1), rep(0,nrow(counts)-nDE))
    dat_alt = create_read_numbers(paramPolyester$mu, paramPolyester$fit,
                                  paramPolyester$p0, m=dim(counts)[1], 
                                  n=sampleSize*2, beta=cbind(log2FC), 
                                  mod=modPolyester)
  }

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(counts),
                               pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = res_alt$baseMean, 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth, 
                               pzero = rowSums(counts(dds_alt)==0)/
                                 ncol(counts(dds_alt))  ))

@@

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 0

if (!file.exists(resfile_n5)){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(null5, file=resfile_n5)
} else {
  null5 <- readRDS(file=resfile_n5)
}
```

@@

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 0

if (!file.exists(resfile_n10)){
  null10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(null10, file=resfile_n10)
} else {
  null10 <- readRDS(file=resfile_n10)
}
```

@@

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 1500

if (!file.exists(resfile_d5)){
  de5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(de5, file=resfile_d5)
} else {
  de5 <- readRDS(file=resfile_d5)
}
```

@@

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 1500

if (!file.exists(resfile_d10)){
  de10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(de10, file=resfile_d10)
} else {
  de10 <- readRDS(file=resfile_d10)
}

@@

>>>>

```{r}
#' @param counts is the full count dataset
#' @param paramPolyester is the parameters from running polyester. 
#' Contains information on the mean-variance relationship used to 
#' simulate RNA-Seq counts. 
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param rseed random seed
simulateOneSplit <- function(X, rseed, counts, paramPolyester, 
                             nDE, sampleSize, bd, uninformativeCovariate = FALSE){
  if(uninformativeCovariate)
    uninf_covar <- rnorm(nrow(counts))

  # set random seed
  set.seed(as.numeric(X)*as.numeric(rseed))

@@

    # randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
    log2FC <- rep(0, nrow(counts))
    log2FC[DE] <- rnorm(nDE, sd = 1)
    dat_alt = create_read_numbers(paramPolyester$mu, paramPolyester$fit,
                                  paramPolyester$p0, m=dim(counts)[1], 
                                  n=sampleSize*2, beta=cbind(log2FC), 
                                  mod=modPolyester)
  }

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(counts),
                               pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = ifelse(uninformativeCovariate, 
                                                      uninf_covar,
                                                      res_alt$baseMean), 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth, 
                               pzero = rowSums(counts(dds_alt)==0)/
                                 ncol(counts(dds_alt))  ))

@@

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 0
rseed <- 225

if (!file.exists(resfile_n5)){
  null5 <- bplapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(null5, file=resfile_n5)
} else {
  null5 <- readRDS(file=resfile_n5)
}
```

@@

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 0
rseed <- 837

if (!file.exists(resfile_n10)){
  null10 <- bplapply(X=1:B, rseed=837, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(null10, file=resfile_n10)
} else {
  null10 <- readRDS(file=resfile_n10)
}
```

@@

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 1500
rseed <- 198

if (!file.exists(resfile_d5)){
  de5 <- bplapply(X=1:B, rseed = rseed, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(de5, file=resfile_d5)
} else {
  de5 <- readRDS(file=resfile_d5)
}
```

@@

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 1500
rseed <- 961

if (!file.exists(resfile_d10)){
  de10 <- bplapply(X=1:B, rseed = rseed, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(de10, file=resfile_d10)
} else {
  de10 <- readRDS(file=resfile_d10)
}

@@

# Uninformative Covariate

## Sim N5: Null Comparison 5v5

### Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 0
rseed <- 225

if (!file.exists(resfile_n5_uninfCov)){
  null5 <- bplapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE)
  saveRDS(null5, file=resfile_n5_uninfCov)
} else {
  null5 <- readRDS(file=resfile_n5_uninfCov)
}
```

### Plot average results over replications

## Sim N10: Null Comparison 10v10

### Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 0
rseed <- 837

if (!file.exists(resfile_n10_uninfCov)){
  null10 <- bplapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE)
  saveRDS(null10, file=resfile_n10_uninfCov)
} else {
  null10 <- readRDS(file=resfile_n10_uninfCov)
}
```

### Plot average results over replications

## Sim D5: DE Comparison 5v5

### Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 1500
rseed <- 198

if (!file.exists(resfile_d5_uninfCov)){
  de5 <- bplapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, counts=counts,
                  paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                  bd=bd, uninformativeCovariate = TRUE)
  saveRDS(de5, file=resfile_d5_uninfCov)
} else {
  de5 <- readRDS(file=resfile_d5_uninfCov)
}
```

### Plot average results over replications

## Sim D10: DE Comparison 10v10

### Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 1500
rseed <- 961

if (!file.exists(resfile_d10_uninfCov)){
  de10 <- bplapply(X=1:B, rseed=rseed, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE)
  saveRDS(de10, file=resfile_d10_uninfCov)
} else {
  de10 <- readRDS(file=resfile_d10_uninfCov)
}

### Plot average results over replications

## Sample size comparison

```{r}
# uninformative covariate
plotMethodRanks(c(resfile_d5_uninfCov, resfile_d10_uninfCov), 
                colLabels = c("DE 5", "DE 10"), 
                alpha = 0.10, xlab = "Comparison",
                excludeMethods = NULL)
```

### 2018-07-17 datasets/RNA-Seq/yeast-simulation.Rmd
Use the same dataset for informative and uninformative covariates within replicate

```{r}
#' @param X simnumber
#' @param seed random seed
#' @param dds is the full DESeq2 dataset
#' @param resFULL is the results object from running DESeq2 on dds (used in
#'   to obtain the fold change effect sizes when adding non-null genes)
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param BPPARAM is the BiocParallel bpparam argument to pass to DESeq. Since
#'  parallelization happens by forking multiple instances of this function, we 
#'  will specify only one worker per DEseq intance.
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param pvalHists logical whether to return histograms of pvalues by covariate
#'  instead of SB objects
simulateOneSplit <- function(X, rseed, dds, resFULL, nDE, sampleSize,
                             bd, uninformativeCovariate = FALSE, pvalHists = FALSE){

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(resTEST), 
                               pval=resTEST$pvalue, 
                               SE=resTEST$lfcSE,                 
                               ind_covariate = resTEST$baseMean,
                               effect_size = resTEST$log2FoldChange, 
                               test_statistic = resTEST$stat,
                               qvalue = truth,
                               pzero = rowSums(counts(dds_test)==0)/
                                       ncol(counts(dds_test)))) %>%
                dplyr::filter(!is.na(pval))

  if(uninformativeCovariate)
    geneExp$ind_covariate <- rnorm(nrow(geneExp))

>>>>

```{r}
#' @param X simnumber
#' @param seed random seed
#' @param dds is the full DESeq2 dataset
#' @param resFULL is the results object from running DESeq2 on dds (used in
#'   to obtain the fold change effect sizes when adding non-null genes)
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param BPPARAM is the BiocParallel bpparam argument to pass to DESeq. Since
#'  parallelization happens by forking multiple instances of this function, we 
#'  will specify only one worker per DEseq intance.
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param pvalHists logical whether to return histograms of pvalues by covariate
#'  instead of SB objects
simulateOneSplit <- function(X, rseed, dds, resFULL, nDE, sampleSize,
                             bd, uninformativeCovariate = FALSE, pvalHists = FALSE){
  if(uninformativeCovariate)
    ind_covar <- rnorm(nrow(dds))

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(resTEST), 
                               pval=resTEST$pvalue, 
                               SE=resTEST$lfcSE,                 
                               ind_covariate = ifelse(uninformativeCovariate, 
                                                      ind_covar,
                                                      resTEST$baseMean),
                               effect_size = resTEST$log2FoldChange, 
                               test_statistic = resTEST$stat,
                               qvalue = truth,
                               pzero = rowSums(counts(dds_test)==0)/
                                       ncol(counts(dds_test)))) %>%
                dplyr::filter(!is.na(pval))
